---
title: "Use dply/ggplot2 to manipulate and explore data"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Load the data and packages
```{r}
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(gapminder))
```

## Task1: Get the maximum and minimum of GDP per capita for all continents.

```{r}
mm=gapminder %>% 
  group_by(continent) %>% 
  summarise(max.gdp=max(gdpPercap), min.gdp=min(gdpPercap))
knitr::kable(mm)
```

Let's visualize this data. Here I will use bar chart to plot maximum and minimum values for each continent seperately. 

```{r minimum gdp}
gapminder %>% 
  group_by(continent) %>% 
  summarise(max=max(gdpPercap), min=min(gdpPercap)) %>%
  gather(key=type, value=GDP.per.capita, 2:3) %>% 
  ggplot(aes(x=continent, y=GDP.per.capita, fill=continent)) + 
  geom_bar(stat = "identity")+
  facet_wrap(.~type, scale="free_y") +
  guides(fill=FALSE)
```


## Task2: Look at the spread of GDP per capita within the continents.

Let's look at the standard deviation of gdp on different continents through `sd()` function
```{r}
stats=gapminder %>% 
  group_by(continent) %>% 
  summarise(spread=sd(gdpPercap),
            min=min(gdpPercap),
            Qu.1st=quantile(gdpPercap, probs= 0.25),
            median=median(gdpPercap),
            mean=mean(gdpPercap),
            Qu.3st=quantile(gdpPercap, probs= 0.25),
            max=max(gdpPercap))
            
knitr::kable(stats)
```


One way to visualize the spread of data is through histogram
```{r spread of GDP per capita within the continents}
gapminder %>% 
  ggplot(aes(gdpPercap))+
  geom_histogram(aes(y=..density.., fill=continent), bins=20)+
  geom_density()+
  facet_wrap(~continent)+
  guides(fill=FALSE)

```

When we want to check the spread of data, using violin along with jitter plot is a good option

```{r spread with violin and jitter plot}
gapminder%>%
  ggplot(aes(continent, gdpPercap, color = continent))+
  geom_jitter(aes(alpha = 0.1))+
  geom_violin() 
  
  
```


## Task3: Compute a trimmed mean or a weighted mean of life expectancy for different years. 

Here I will explore the mean weighting by population
```{r}
df_m=gapminder %>% 
  group_by(year) %>% 
  summarise(weighted_lifeExp=weighted.mean(lifeExp, pop))
knitr::kable(df_m)
```

Let's visualize this data
```{r weighted mean lifeExp over time}
gapminder %>% 
  group_by(year) %>% 
  summarise(weighted_mean=weighted.mean(lifeExp, pop)) %>% 
  ggplot(aes(year, weighted_mean)) + 
  geom_point()+
  geom_smooth(method = "lm")
```



## Task4: How is life expectancy changing over time on different continents?
Here, I will use weighted mean of lifeExp to illustrate its change over time
```{r}
life=gapminder %>%
  group_by(continent, year) %>%
  summarise(weighted_lifeExp=weighted.mean(lifeExp, pop))
knitr::kable(life)
```

```{r lifeExp vs. time on different continents}
life %>% 
  ggplot(aes(year, weighted_lifeExp, color=continent)) + 
  geom_point()+
  geom_smooth(method = "lm")

```

## Task5: Report the absolute and/or relative abundance of countries with low life expectancy over time by continent: 

Here, I will first compute the mean value of worldwide life expectancy, then use this value as a benchmark to determine how many countries on each continent have a life expectancy less than this value, for each year.

```{r}
mean=summarise(gapminder,mean(lifeExp)) ## compute mean worldwide life expectancy
## knowing the mean value of worldwide life expectancy is 59.5, I will count how many countries within each continent have a ife expectancy less than this value
df_c=gapminder %>% 
  group_by(continent) %>% 
  filter(lifeExp<59.5) %>% 
  tally()
knitr::kable(df_c)
```

From this table we can see that Africa and Asia countries have more population have lower life expectancy compared to countries on other continents.


Now let's plot the data. I will first plot lifeExp of all range for all continents but use mean value as a benchmark to differentiate life expectancy above and below such value.
```{r lifeExp greater than 59.5}
gapminder %>% 
  ggplot(aes(year, lifeExp)) + 
  geom_point(aes(color=lifeExp>=59.5))+ # use mean lifeExp as a benchmark 
  facet_wrap(~continent) +
  geom_smooth()
```



